---
title: "Import Onset HOBO Logger Data into R"
author: "Matthew Van Scoyoc, National Park Service"
date: "5 April, 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{raindancer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Introduction
This R package imports data from Onset temperature, relative humidity, and precipitation (event) data loggers into R. 
Data collected in the field using Onset loggers are exported to comma delimited (csv) files using the HOBOware application from Onset.
This package imports the csv files into R and summarizes the data.

The structure of the csv files generated from HOBOware vary wildly. 
They can have anywhere from 4 to 10 columns and the logger details are usually in two "hidden" columns following the data. 
This makes data from the temperature, relative humidity, and precipitation loggers difficult to read into R. 
This package imports data collected by Onset loggers used in the Southeast Utah Group (SEUG) national parks long-term vegetation monitoring program (LTVMP) and then summarize these data.
This package was written to be used with the dataprocessR package, [https://github.com/scoyoc/dataprocessR](https://github.com/scoyoc/dataprocessR), that exports the raw and summarized data to the SEUG LTVMP database (a Microsoft Access database).

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library("raindancer")
```

# Installation
This package is available on [GitHub](https://github.com/) at [https://github.com/scoyoc/raindancer](https://github.com/scoyoc/raindancer). 
Dependent packages include dplyr, glue, lubridate, RODBC, stringr, tibble, tidyr, and utils. 
Suggested pacakges include janitor, knitr, rmarkdown, and readr. 
```{r Setup, eval=FALSE, echo = TRUE}
if (!"devtools" %in% installed.packages()[, "Package"]) {
  install.packages("devtools")
}
devtools::install_github("scoyoc/raindancer")
library("raindancer")
```

# Import Data into R
There are two function available to import the csv files generated from HOBOware into R, *import_hobo_2008()* and *import_hobo()*. Below is a table of Onset loggers that these functions can import data from.
```{r logger_table, eval=TRUE, echo=FALSE}
onset_loggers <- data.frame(
  'Product' = c("H07 Logger", "HOBO UA-003-64 Pendant Temp/Event",
                "H08 Logger", "HOBO UA-001-64 Pendant Temp", 
                "HOBO U23-001 Temp/RH"),
  'Element' = c("PRCP", "PRCP & TEMP", "TEMP", "TEMP", 
                "TEMP & RH"),
  'Years' = c("2008-2019", "2019-present", "2008-2019", 
                   "2019-present", "2019-present")
)

knitr::kable(onset_loggers, 
             col.names = c("Onset Product", "Element Recorded", 
                           "Years Used"), 
             caption = "Table. Onset loggers used in the SEUG LTVMP.",
             format = 'html')
```

The functions *import_hobo_2008()* and *import_hobo()* return a list with three components:  
1. **file_info** is a one row data frame that contains information about the file, the logger, and data.  
2. **details** is a data frame of logger and sampling event information.  
3. **data_raw** is a data frame of raw data.

Lets start by usgin *list.files()* to bring a list of csv files into R. 
```{r import_example, eval=FALSE, echo=TRUE}
file_list <- list.files(path = "C:/path/to/data", pattern = ".csv", 
                        full.names = TRUE, recursive = FALSE)
```

There are some files included in this package for examples, so we'll use those for the vignette. 
```{r import, eval=TRUE, echo=TRUE, results='markup'}
file_list <- list.files(path = system.file("extdata", package = "raindancer"),
                        pattern = ".csv", full.names = TRUE, recursive = FALSE)
print(file_list)
```

The first four csv files are from 2010 and 2012, so lets start by using *import_hobo_2008()*.
```{r import_hobo_2008.R, eval=TRUE, echo=TRUE, results='markup'}
dat.1 <- import_hobo_2008(file_list[1])
```

Let's examine the components of the list returned by the *import_hobo_2008()* function.
The first component returns information about the file, the logger, and the data.
```{r file_info, eval=TRUE, echo=TRUE, results='markup'}
dplyr::glimpse(dat.1$file_info)
```

The second component returns a three (3) column data frame with information about the logger and sampling event.
```{r details, eval=TRUE, echo=TRUE, results='markup'}
dat.1$details
```

And the third component returns a data frame of raw data.
```{r data_raw, eval=TRUE, echo=TRUE, results='markup'}
dplyr::glimpse(dat.1$data_raw)
```

In 2019 SEUG resource staff upgraded the loggers in hopes of preventing logger failure and mainaining the weather data set for the LTVMP.
The new loggers had a different file structure that the older loggers, requiring a new function to import data into R. 
Files five (5) though eight (8) are examples from the new loggers. 
Let's use *import_hobo()* to bring data from one of these files into R and examine the information and data.
```{r import_hobo.R, eval=TRUE, echo=TRUE, results='markup'}
dat.6 <- import_hobo(file_list[6])
str(dat.6)
```

# Summarize Data





